Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome

### 2017-08-08 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A putative causal map of natural variation across the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{A putative causal map of natural variation across the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 August 2017}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a
range of pleiotropy models, instrument selection will be increasingly
liable to selecting invalid instruments as GWAS sample sizes continue to
grow. Methods have been developed in an attempt to protect MR from
different patterns of horizontal pleiotropy, and here we have designed a
mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts
the most appropriate model to use for any specific causal analysis,
improving on both power and false discovery rates. Using the approach,
we systematically estimated the causal effects amongst 2407 phenotypes,
comprising the evaluation of causality between over a million pairs of
traits. Almost 90\% of causal estimates indicated some level of
horizontal pleiotropy. The causal estimates are organised into a
publicly available graph database (\url{http://eve.mrbase.org}), and we
use it here to highlight the numerous challenges that remain in
automated causal inference.

@@

Genome-wide association studies (GWAS) have identified potential
instrumental variables for thousands of phenotypes (10). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using either individual-level or only summary-level data
(known as two-sample MR, 2SMR) (11). Here, in order to infer the causal
effect of an exposure on an outcome all that is required are the genetic
effects of the instrumenting SNP on the exposure and the outcome. This
has three major advantages. First, GWAS summary data are non-disclosive
and often publicly available. Second, causal inference can be made
between phenotypes even if they have not been measured in the same
samples, limiting possible MR analyses only by the availability of GWAS
summary data for the traits in question (12). Third, because each
instrumental variable mimics an independent randomization study, we can
view the causal inference problem through the simple and widely
understood prism of a meta-analysis.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure (Box 1).
The extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (4--7,9,13). It is considered best
practice to triangulate estimates from a range of MR methods (and other
experimental designs) when presenting causal estimates so as to assess
the potential for different horizontal pleiotropy models and other
violations of MR assumptions (14). However there are circumstances in
which it is desirable to consolidate across many methods to obtain a
single MR estimate. First, it could be of interest to screen links
between thousands of traits for being potential causal associations, in
which case a critical evaluation of each causal estimate of interest may
not be possible or convenient. Second, though the simplest method, based
on inverse-variance weighted (IVW) meta analysis (11), is most
statistically powerful under no HP, it can have high false positive or
low true positive rates in the presence of HP compared to other methods.
Pleiotropy has been hypothesised to be universal (15), though the degree
and nature (e.g.~horizontal versus vertical, see Box 1) to which this
may be the case is contested (16,17), hence defaulting to the IVW method
in the first instance and using other methods as sensitivity analyses
may not be appropriate. Third, if different methods disagree it is not
possible to know which is correct because the true nature of HP
exhibited by the instruments is not known. Fourth, the available methods
do not cover all possible models of HP, and therefore an automated
method for instrument selection may be necessary.

@@

\textbf{Mean-based methods:} Here we consider four nested models (6).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression line
without affecting the slope (and therefore not introducing bias). Fixed
effects Egger regression (4) relaxes the HP assumption further by
allowing a non-zero intercept which essentially allows overall
horizontal pleiotropy to be directional, where its total effect
influences the outcome in a specific direction. Random effects Egger
regression further allows heterogeneity around the slope having
accounted for overall directional HP (6), as long as the HP effects are
not correlated with the SNP-exposure effects (also known as the INSIDE
assumption)(4).

@@

The simplest approach to selecting instruments is to take SNPs that have
been declared significant in the published GWAS for the exposure. This
typically involves obtaining SNPs that surpass \(p < 5 \times 10^{-8}\),
using clumping to obtain independent SNPs, and then replicating in an
independent sample. These results are often recorded in public GWAS
catalogs. Alternatively the clumping procedure can be performed using
complete summary data in MR-Base (12). Complete summary data refers to
association results for all SNPs used in a GWAS i.e.~not only those
passing statistical thresholds for significance. We call this the
``tophits'' strategy.

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously (8). However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid. Steiger's Z-test of correlated correlations (31) can be
used to formally test the extent to which the two correlations are
statistically different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). Other methods have been developed to identify
invalid instruments for the purposes of exclusion from MR analyses
(9,13,32), based on the notion that outlying instruments are more likely
to be a source of HP. A potential drawback of this approach is that the
outlier SNPs in a summary-set might be the only reliable ones. This
could arise, for example, for the \emph{CRP} variant instrumenting
C-reactive protein levels (21), or the \emph{SLC2A9} variant influencing
serum urate levels (33).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total (Supplementary table 2). Each method assumes or
performs best for a different model of pleiotropy. Our objective is to
select the method most likely to be correct for a specific MR analysis
by predicting the model of pleiotropy, and relating that prediction to
the most appropriate model. To achieve this we use a ``mixture of
experts'' machine learning approach (34), where each MR method is
considered to be an `expert', taking a summary-set as its input. An
overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts
approach of (34) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (35--37) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR methods are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If the number of variants after Steiger filtering is greater than 5
  then apply the MoE to obtain the best method. This value is chosen
  arbitrarily as a minimum number of variants for which the different MR
  methods can model HP
\item
  If the number of variants after Steiger filtering is less than or
  equal to 5 but greater than 1 then use the IVW random effects approach
  on the filtered set of variants
\item
  If there is 1 variant retained in the Steiger approach then use the
  Wald ratio on the remaining variant
\item
  If there are no variants remaining after Steiger filtering then
  declare no estimate of a causal association possible.
\end{enumerate}

@@

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered for an outcome trait that acts
primarily on some exposure trait for that outcome increases (Figure 1a).
We evaluated the efficacy of the Steiger filtering approach for
improving instrument selection using 100,000 simulated summary-sets
comprising both null and non-null causal models. For each summary-set,
instruments were selected based on the tophits strategy and the Steiger
filtering strategy, and MR was performed using 14 different methods
based on instruments selected from each of these strategies.

Figure 1b shows that the tophits strategy led to over half of the
instruments being primarily associated with either confounders or the
outcome phenotype, not the exposure phenotype. For brevity, we refer to
the latter type of invalid instruments as ``reverse causal''
instruments. The proportion of invalid instruments due to reverse cause
increased as GWAS discovery sample size increased. Applying Steiger
filtering reduced this to 25\%. Consequently, the false discovery rates
(FDR) for 12 of the 14 methods reduced substantially when applied using
Steiger filtered instruments (Figure 1c). The true positive rates for
the methods based on Steiger filtering did however reduce slightly for
10 of the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR methods
and instrument selection strategies to apply to a particular summary-set
based on its characteristics (Figure 2).

@@

We compared the MoE performance in the simulations against each of the
28 methods, testing to see if it outperformed all other single methods.
Figures 3a and 3b show that the MoE approach had the best general
performance. Estimating the area under the receiver operator curve gave
0.84 for the MoE approach in terms of classifying the simulations as
being null or non-null. Notably, the next best methods were median and
mode based estimators using Steiger filtering. But a crucial observation
is that under the assumption of widespread and diverse pleiotropic
effects it is clear that all methods suffer from high false discovery
rates (Figure 3a), including the MoE approach.

\subsubsection{Automated MR analysis of 2407
phenotypes}\label{automated-mr-analysis-of-2407-phenotypes}

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown
in Supplementary table 3. Amongst those deemed `significant', the IVW
fixed effects analysis method combined with the tophits instruments
selection strategy (only valid when there is no detectable evidence of
horizontal pleiotropy of any sort) was selected in only 10.4\% of cases.
This indicates that horizontal pleiotropy is likely to be pervasive.

\subsubsection{Putative associations involving LDL cholesterol and years
of
schooling}\label{putative-associations-involving-ldl-cholesterol-and-years-of-schooling}

As an example of the database we performed a look up of associations
between all traits and LDL cholesterol (as measured in the GLGC
consortium (41)), where the MoE result had a false discovery rate of
0.05 using the MoE approach. This returned 287 putative associations,
amongst which 111 involved LDL cholesterol influencing other traits and
176 involved other traits influencing LDL cholesterol. A large
proportion of these traits were metabolites, which are dominated by
lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
returned 27 associations with 23 traits (Figure 4a). There was a strong
association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
after Steiger filtering, indicating the presence of HP being present
amongst the instruments.

@@

The facility to perform these scans has been automated in the MR of
everything vs everything (MR-EvE) web application,
\url{http://eve.mrbase.org}.

@@

Methods have been introduced recently that seek to improve MR estimates
through invalid instrument detection and removal (9,13,32). These attack
the problem from the perspective that variants which have a relatively
large influence on the causal estimate, or provide substantially
different estimates from the other variants, are likely to be exhibiting
HP, and their removal from the analysis may be warranted. A potential
drawback of this approach is that the outlier SNPs in a summary-set
might be the only reliable ones. This could arise, for example, for the
\emph{CRP} variant instrumenting C-reactive protein levels (21), because
CRP itself is downstream of many cellular functions and most GWAS
signals will not be specific for the trait. Steiger filtering has
attempted to sidestep this issue in removing invalid instruments,
seeking to detect those instruments which have associations with the
outcome that are too large to be consistent with the exposure causing
the outcome. We acknowledge that unmeasured confounding and measurement
error can introduce problems with Steiger filtering (8), but note that
our simulations indicate that all individual MR methods improve in
performance following its application.

In addition to HP, there are many other limitations that prevent MR from
being a panacea for causal inference. Many of the associations are
biologically impossible, for example where early life phenotypes appear
to be influenced by later stage phenotypes. Such associations can be
explained statistically because the instruments proxy the liability for
that particular phenotype, not that phenotype's observed manifestation.
Though these associations can be informative, their interpretations as
causal relationships are far from clear. Similarly, often disease traits
appear to causally relate to other phenotypes, but GWAS of disease is
typically performed on the liability scale, hence the causal estimate
reflects not the presence or absence of disease, but the underlying risk
of disease. Again, interpretation of such associations can be
problematic due to the instruments proxying disease liability and not
diseae outcomes.

@@

Other problems, in addition to HP, can also manifest. Frailty effects
(48) could induce associations for late-onset traits, and often detailed
modelling for how this can influence MR is warranted (43,49). Genetic
variants could strongly relate to several phenotypes, for example SNPs
that influence LDL cholesterol are also likely to influence other lipid
fractions, making it difficult to ascertain which amongst them is the
true causal exposure (50). We see this issue manifesting in Figure 4,
where there appear to be causal relationships between different lipid
fractions. The measured feature for which genetic associations are known
could relate in complicated ways to the biological entities that is
truly causal (2), for example higher circulating levels of the natural
antioxidant extracellular superoxide dismutase (EC-SOD) appears to
causally relate to higher risk of coronary heart disease (51) because
the phenotypic measure relates inversely to the EC-SOD levels in
arterial walls, hence \emph{in situ} activity is lower (2). Dynastic
effects can also confound MR estimates, arising when the exposure trait
in a previous generation influences the environment of the current
generation. Here, the SNPs being used as instruments for the exposure
will be associated with a potential confounder in the analysis (52). The
associations depicted in Figure 4 involving educational attainment may
be susceptible to bias arising from this mechanism, and protecting
against it will likely require alternative study designs, for example by
using sibling pairs.

We also encounter a potential new problem in using machine learning
approaches to infer the correct model for a particular summary-set,
namely, that the model selection is data driven. This is liable to
contribute to elevated type 1 error rates, even though we have attempted
to separate the information used for optimisation from the information
used to predict performance. One advantage to generating a method that
consolidates across many methods is that it can prevent cherry picking
the most appealing method based on their results. The extent to which a
machine-based method selector incurs higher type 1 error rates more than
a human selector is not clear. Though MR-MoE does exhibit higher type 1
error rates than are desirable, they still remain amongst the lowest
compared to other MR methods that do not suffer from the issue of data
driven model selection.

@@

Figure 3: Performance of MoE against all other methods. a) The power for
non-null datasets is plotted against the FDR for null datasets for each
of the 28 methods, plus MoE. No single method achieved nominal FDR for
these simulations. b) Calculating the area under the ROC curve from the
values in (a) we plotted the performance in order from lowest to
highest. Under the assumption of pervasive horizontal pleiotropy, the
MoE approach is likely most effective than any other single method.

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Automating Mendelian randomization through machine learning to construct a putative causal map of the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Automating Mendelian randomization through machine learning to construct
a putative causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{08 August 2017}

@@

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomization
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy. We demonstrate that across a
range of pleiotropy models, instrument selection will be increasingly
liable to selecting invalid instruments as GWAS sample sizes continue to
grow. Methods have been developed in an attempt to protect MR from
different patterns of horizontal pleiotropy, and here we have designed a
mixture-of-experts machine learning framework (MR-MoE 1.0) that predicts
the most appropriate model to use for any specific causal analysis,
improving on both power and false discovery rates. Using the approach,
we systematically estimated the causal effects amongst 2407 phenotypes.
Almost 90\% of causal estimates indicated some level of horizontal
pleiotropy. The causal estimates are organised into a publicly available
graph database (\url{http://eve.mrbase.org}), and we use it here to
highlight the numerous challenges that remain in automated causal
inference.

@@

Genome-wide association studies (GWAS) have identified potential
instrumental variables for thousands of phenotypes (10). Recent
developments in MR have enabled knowledge of instrumental variables to
be applied using either individual-level or only summary-level data
(known as two-sample MR, 2SMR) (11). Here, in order to infer the causal
effect of an exposure on an outcome all that is required are the genetic
effects of the instrumenting SNP on the exposure and the outcome. This
has three major advantages. First, GWAS summary data are non-disclosive
and often publicly available. Second, causal inference can be made
between phenotypes even if they have not been measured in the same
samples, limiting possible MR analyses only by the availability of GWAS
summary data for the traits in question (12). Third, because each
instrumental variable mimics an independent randomization study, we can
view the causal inference problem through the simple and widely
understood prism of a meta-analysis (6).

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy, where they
influence the outcome through a pathway other than the exposure (Box 1).
The extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of horizontal pleiotropy (4--7,9,13). It
is considered best practice to triangulate estimates from a range of MR
methods (and other experimental designs) when presenting causal
estimates so as to assess the potential for different horizontal
pleiotropy models and other violations of MR assumptions (14). However
there are circumstances in which it is desirable to consolidate across
many methods to obtain a single MR estimate. First, it could be of
interest to screen links between thousands of traits for being potential
causal associations, in which case a critical evaluation of each causal
estimate of interest may not be possible or convenient. Second, though
the simplest method, based on inverse-variance weighted (IVW) meta
analysis (11), is most statistically powerful under no horizontal
pleiotropy, it can have high false positive or low true positive rates
in the presence of horizontal pleiotropy compared to other methods.
Pleiotropy has been hypothesised to be universal (15), though the degree
and nature (e.g.~horizontal versus vertical, see Box 1) to which this
may be the case is contested (16,17), hence defaulting to the IVW method
in the first instance and using other methods as sensitivity analyses
may not be appropriate. Third, if different methods disagree it is not
possible to know which is correct because the true nature of horizontal
pleiotropy exhibited by the instruments is not known. Fourth, the
available methods do not cover all possible models of horizontal
pleiotropy, and therefore an automated method for instrument selection
may be necessary.

@@

\textbf{Mean-based methods:} Here we consider four nested models (6).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no horizontal pleiotropy. IVW random
effects meta-analysis relaxes the horizontal pleiotropy assumption,
allowing it to be present but balanced - such that it only leads to
increased heterogeneity around the regression line without affecting the
slope (and therefore not introducing bias). Fixed effects Egger
regression (4) relaxes the horizontal pleiotropy assumption further by
allowing a non-zero intercept which essentially allows overall
horizontal pleiotropy to be directional, where its total effect
influences the outcome in a specific direction. Random effects Egger
regression further allows heterogeneity around the slope having
accounted for overall directional horizontal pleiotropy (6), as long as
the horizontal pleiotropy effects are not correlated with the
SNP-exposure effects (also known as the INSIDE assumption)(4).

@@

The simplest approach to selecting instruments is to take SNPs that have
been declared significant in the published GWAS for the exposure. This
typically involves obtaining SNPs that surpass \(p < 5 \times 10^{-8}\),
using clumping to obtain independent SNPs, and then replicating in an
independent sample. These results are often recorded in public GWAS
catalogs. Alternatively the clumping procedure can be performed using
complete summary data in MR-Base (12). Complete summary data refers to
association results for all SNPs used in a GWAS i.e.~not only those
passing statistical thresholds for significance. We call this the
``top-hits'' strategy.

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(cor(x, x_o) < cor(x,y)cor(y,y_o)\) where \(cor(x, x_o)\) and
\(cor(y, y_o)\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(cor(g,x)\) and \(cor(g,y)\) parameters, as has been discussed in
detail previously (8). Large differences in sample sizes between the
exposure and outcome GWASs may also have a practical impact on the
efficacy of this approach. However, we use it here as a computationally
inexpensive and approximate method to identify variants that are likely
to be invalid (Figure 1a). Steiger's Z-test of correlated correlations
(31) can be used to formally test the extent to which the two
correlations are statistically different.

Other methods have been developed to identify invalid instruments for
the purposes of exclusion from MR analyses (9,13,32), based on the
notion that outlying instruments are more likely to be a source of
horizontal pleiotropy. A potential drawback of this approach is that the
outlier SNPs in a summary-set might be the only reliable ones. This
could arise, for example, for the \emph{CRP} variant instrumenting
C-reactive protein levels (21), or the \emph{SLC2A9} variant influencing
serum urate levels (33).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
strategies in total (Supplementary table 2). Each method assumes or
performs best for a different model of pleiotropy. Our objective is to
select the method most likely to be correct for a specific MR analysis
by predicting the model of pleiotropy, and relating that prediction to
the most appropriate model. To achieve this we use a ``mixture of
experts'' machine learning approach (34), where each MR strategy is
considered to be an `expert', taking a summary-set as its input. An
overview of the approach is shown in Figure 2.

@@

It should be noted that in the original hierarchical mixture of experts
approach of (34) the gating function and the experts share the same
input space and are trained simultaneously using
expectation-maximisation. In our approach the gating function is defined
over a separate input space consisting of the parameter space of the
experts, and is trained separately in a supervised setting using
simulated data. As such our MR-MoE approach fits the framework of
meta-learning (35--37) where machine learning methods are applied at the
`meta-level' to analyse the results of machine learning experiments at
the `base-level', in order to be able to understand and model the
capabilities of each expert and recommend which expert should be applied
to a given problem. From this perspective the 28 MR strategies are the
base-models, the gating function is the meta-model, and its
meta-features are the parameters of the summary-sets.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  If the number of variants after Steiger filtering is greater than 5
  then apply the MoE to obtain the best method. This value is chosen
  arbitrarily as a minimum number of variants for which the different MR
  methods can model horizontal pleiotropy
\item
  If the number of variants after Steiger filtering is less than or
  equal to 5 but greater than 1 then use the IVW random effects approach
  on the filtered set of variants
\item
  If there is 1 variant retained in the Steiger approach then use the
  Wald ratio on the remaining variant
\item
  If there are no variants remaining after Steiger filtering then
  declare no estimate of a causal association possible.
\end{enumerate}

@@

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered for a trait that acts primarily
on one of its precursors increases (Figure 1a). This presents a problem
when GWAS significance is used as the sole criterion for instrument
selection if the hypothesis being tested is either a) the trait causing
the precursor, or b) the trait causing some other trait and the
precursor causally relates to both (in which case the precursor is a
confounder). We evaluated the efficacy of the Steiger filtering approach
for improving instrument selection using 100,000 simulated summary-sets
comprising both null and non-null causal models. For each summary-set,
instruments were selected based on the top-hits strategy and the Steiger
filtering strategy, and MR was performed using 14 different methods
based on instruments selected from each of these strategies.

Figure 1b shows that the top-hits strategy led to over half of the
instruments being primarily associated with either confounders or the
outcome phenotype, not the exposure phenotype. For brevity, we refer to
the latter type of invalid instruments as ``reverse causal''
instruments. The proportion of invalid instruments due to reverse cause
increased as GWAS discovery sample size increased. Applying Steiger
filtering reduced this to 25\%. Consequently, the false discovery rates
(FDR) for 12 of the 14 methods reduced substantially when applied using
Steiger filtered instruments (Figure 1c). The true positive rates for
the methods based on Steiger filtering did however reduce slightly for
10 of the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to predict the most appropriate of the 28 MR
strategies and instrument selection strategies to apply to a particular
summary-set based on its characteristics (Figure 2).

@@

We compared the MoE performance in the simulations against each of the
28 strategies, testing to see if it outperformed all other single
strategies. Figures 3a and 3b show that the MoE approach had the best
general performance. Estimating the area under the receiver operator
curve gave 0.84 for the MoE approach in terms of classifying the
simulations as being null or non-null. Notably, the next best methods
were median and mode based estimators using Steiger filtering. But a
crucial observation is that under the assumption of widespread and
diverse pleiotropic effects it is clear that all methods suffer from
high false discovery rates (Figure 3a), including the MoE approach.

\subsubsection{Automated MR analysis of 2407 phenotypes reveals
substantial horizontal
pleiotropy}\label{automated-mr-analysis-of-2407-phenotypes-reveals-substantial-horizontal-pleiotropy}

@@

The frequencies of the methods chosen by the MR-MoE analysis are shown
in Supplementary table 3. Amongst those deemed `significant', the IVW
fixed effects analysis method combined with the top-hits instruments
selection strategy (only valid when there is no detectable evidence of
horizontal pleiotropy of any sort) was selected in only 10.4\% of cases.
This indicates that horizontal pleiotropy is likely to be pervasive.

\subsubsection{Interpretation of the putative associations in a broad
scan of phenotypes requires detailed follow
up}\label{interpretation-of-the-putative-associations-in-a-broad-scan-of-phenotypes-requires-detailed-follow-up}

As an example of the database we performed a look up of associations
between all traits and LDL cholesterol (as measured in the GLGC
consortium (41)), where the MoE result had a false discovery rate of
0.05 using the MoE approach. This returned 287 putative associations,
amongst which 111 involved LDL cholesterol influencing other traits and
176 involved other traits influencing LDL cholesterol. A large
proportion of these traits were metabolites, which are dominated by
lipid fractions. Filtering to exclude the metabolomic studies ((38,39))
returned 27 associations with 23 traits (Figure 4a). There was a strong
association with coronary heart disease (0.41 log(OR) per SD, 0.31-0.52,
\(p=2.7\times10^{-12}\)). Here the IVW random effects method was chosen
after Steiger filtering, indicating the presence of horizontal
pleiotropy being present amongst the instruments.

@@

The facility to perform these scans has been automated in the MR of
`everything vs everything' (MR-EvE) web application,
\url{http://eve.mrbase.org}.

@@

Methods have been introduced recently that seek to improve MR estimates
through invalid instrument detection and removal (9,13,32). These attack
the problem from the perspective that variants which have a relatively
large influence on the causal estimate, or provide substantially
different estimates from the other variants, are likely to be exhibiting
horizontal pleiotropy, and their removal from the analysis may be
warranted. A potential drawback of this approach is that the outlier
SNPs in a summary-set might be the only reliable ones. This could arise,
for example, for the \emph{CRP} variant instrumenting C-reactive protein
levels (21), because CRP itself is downstream of many cellular functions
and most GWAS signals will not be specific for the trait. Steiger
filtering has attempted to sidestep this issue in removing invalid
instruments, seeking to detect those instruments which have associations
with the outcome that are too large to be consistent with the exposure
causing the outcome. We acknowledge that unmeasured confounding and
measurement error can introduce problems with Steiger filtering (8), but
note that our simulations indicate that all individual MR methods
improve in performance following its application.

In addition to horizontal pleiotropy, there are many other limitations
that prevent MR from being a panacea for causal inference. Many of the
associations are biologically impossible, for example where early life
phenotypes appear to be influenced by later stage phenotypes. Such
associations can be explained statistically because the instruments
proxy the liability for that particular phenotype, not that phenotype's
observed manifestation. Though these associations can be informative,
their interpretations as causal relationships are far from clear.
Similarly, often disease traits appear to causally relate to other
phenotypes, but GWAS of disease is typically performed on the liability
scale, hence the causal estimate reflects not the presence or absence of
disease, but the underlying risk of disease. Again, interpretation of
such associations can be problematic due to the instruments proxying
disease liability and not diseae outcomes.

@@

Other problems, in addition to horizontal pleiotropy, can also manifest.
Frailty effects (48) could induce associations for late-onset traits,
and often detailed modelling for how this can influence MR is warranted
(43,49). Genetic variants could strongly relate to several phenotypes,
for example SNPs that influence LDL cholesterol are also likely to
influence other lipid fractions, making it difficult to ascertain which
amongst them is the true causal exposure (50). We see this issue
manifesting in Figure 4, where there appear to be causal relationships
between different lipid fractions. The measured feature for which
genetic associations are known could relate in complicated ways to the
biological entities that is truly causal (2), for example higher
circulating levels of the natural antioxidant extracellular superoxide
dismutase (EC-SOD) appears to causally relate to higher risk of coronary
heart disease (51) because the phenotypic measure relates inversely to
the EC-SOD levels in arterial walls, hence \emph{in situ} activity is
lower (2). Dynastic effects can also confound MR estimates, arising when
the exposure trait in a previous generation influences the environment
of the current generation. Here, the SNPs being used as instruments for
the exposure will be associated with a potential confounder in the
analysis (52). The associations depicted in Figure 4 involving
educational attainment may be susceptible to bias arising from this
mechanism, and protecting against it will likely require alternative
study designs, for example by using sibling pairs.

A new problem may arise when using machine learning approaches to infer
the correct model for a particular summary-set, namely, that the model
selection is data driven. This is liable to contribute to elevated type
1 error rates, even though we have attempted to separate the information
used for optimisation from the information used to predict performance.
One advantage to generating a method that consolidates across many
methods is that it can prevent cherry picking the most appealing method
based on their results. The extent to which a machine-based method
selector incurs higher type 1 error rates more than a human selector is
not clear. Though MR-MoE does exhibit higher type 1 error rates than are
desirable, they still remain amongst the lowest compared to other MR
methods that do not suffer from the issue of data driven model
selection.

@@

Figure 3: Performance of MoE against all other MR strategies. a) The
power for non-null datasets is plotted against the FDR for null datasets
for each of the 28 strategies, plus MoE. No single MR strategy achieved
nominal FDR for these simulations. b) Calculating the area under the ROC
curve from the values in (a) we plotted the performance in order from
lowest to highest. Under the assumption of pervasive horizontal
pleiotropy, the MoE approach is likely most effective than any other
single MR strategy.
